---
title: "Capital Punishment in the U.S."
subtitle: "STT2860 Project 02 Fall 2022"
author: "GROUP MEMBER NAMES HERE"
date: "`r format(Sys.time(), '%A, %B %d, %Y @ %I:%M %p')`"
output: 
  html_document: 
    theme: yeti
    highlight: textmate
    toc: TRUE
    toc_depth: 4
---
<hr>

```{r globaloptions, include = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment = NA
)
```

### Packages Used

```{r}
library(tidyverse)
```

### Read Data Files

```{r}
cappunish <- readRDS("cappunish.rds")
deathrowsize <- readRDS("deathrowsize.rds")
deathsentencesNC <- readRDS("deathsentencesNC.rds")
```

### Plot #1: How Many States Have Capital Punishment? 

* Create a barplot for the variable `dp2`. 
* Use `color =` and `fill =` to format the bars.
* Use `labs()` to add a title and better axis labels.
* Apply a built-in theme from **ggplot2** or **ggthemes**.
* You may optionally add a subtitle or caption to the plot.
* This plot should *not* have a visible legend anywhere on it.

```{r}
ggplot(cappunish, aes(x = dp2)) +
	geom_bar(color = "black", fill = "skyblue") +
	labs(
    title = "Death Penalty Options",
    x = "Legality of Death Penality, or Moratorium",
    y = "Number of States") +
	theme_bw() 

```

### Plot #2: Capital Punishment By Region of the U.S.

* Create a barplot for the variable `region`. 
* Use `fill =` to shade each bar according to `dp2`.
* Change the `position =` to set each bar equal to 100%.
* Use `width =` to make the bars narrower than the default.
* Use `labs()` to add a title, axis labels, and a better legend label.
* Use `scale_fill_manual()` to choose a different color scheme for the plot. 
* Use `theme()` to move the legend to the bottom of the plot instead of the right
* You may optionally add a subtitle, caption, or ggplot2 built-in theme to the plot.

```{r}
ggplot(cappunish, aes(x = region, fill = dp2)) +
	geom_bar(width = 0.5, position = "fill") +
	labs(
    title = "Death Penalty Options",
    x = "Legality of Death Penality, or Moratorium",
    y = "Percentage of States") +
	scale_fill_manual(values = c("lightsalmon", "00006", "00996")) +
	theme(legend.position = "bottom" )
```

### Plot #3: Number of Prisoners on Death Row

* Create a barplot for the total number of prisoners (y) in each year (x).
* Fill the bars using the variable `decade` and outline them all in black.
* Set the width of all the bars so there is no visible gap in between them.
* Scale the x-axis to include a tick mark at each decade (1970, 1980, etc).
* Scale the y axis so that it is numbered 0, 500, 1000, 1500, 2000, etc.
* Use `labs()` to add or change the title, axis labels, and legend label. 
* Use `theme()` to do the following:
    * center your title
    * set the plot background to white or blank
    * set all axis tick marks to transparent or blank
    * set the major x-axis gridlines to transparent or blank
    * set the major y-axis gridlines to light gray
* You may optionally change the bar color scheme used in your plot

```{r}
ggplot(deathrowsize, aes(x = year, y = total, fill = decade)) +
	geom_bar(stat = 'identity', color = "black", width = .9) +
	scale_x_continuous(breaks = seq(1960, 2020, 10)) + 
	scale_y_continuous(breaks = seq(0, 4000, 500)) +
	labs(title = "Number of Prisoners on Death Row",
    x = "Year",
    y = "Total Number of Prisoners",
    fill = "Decades") +
	theme(
    plot.title = element_text(hjust = 0.5),
     panel.background = element_blank(),
     axis.ticks.x = element_blank(),
     axis.ticks.y = element_blank(),
     panel.grid.major.x = element_line(colour = "transparent"),
     panel.grid.major.y = element_line(colour = "lightgray")
    )
```

### Plot #4: Number of Executions Since 1976

* Create a violin plot for the variable `post1976` (use `y = 0`)
* Add a boxplot and set the width so it fits inside the violin plot
* Fill each plot with a color of your choice (be colorblind-aware)
* Scale the x-axis of the plot so that it counts in increments of 25
* Get rid of all y-axis markings/gridlines using `scale_y_continuous()`
* Add a title, x-axis label, and a caption to identify the large outlier
* Add the built-in theme `theme_linedraw()` or `theme_minimal` to the plot

Hint: You can use the `NULL` vector in functions like `scale_y_continuous()`

```{r}
ggplot(cappunish, aes(post1976, y = 0)) +
	geom_violin(fill = "lightblue") +
	geom_boxplot(width = 0.1, fill = "yellow") +
	scale_x_continuous(breaks = seq(0, 600, 25)) +
	scale_y_continuous(breaks = NULL) +
	labs(
    title = "Number of Executions Since 1976",
    x = "Number of Excecutions",
    caption = "The extreme outlier is Texas") +
	theme_linedraw() +
    theme(axis.title.y = element_blank()
    )
```

### Plot #5: Death Sentences in NC Since 1976

Create a time series plot of death sentences in North Carolina. Use three geometries on this plot: `geom_area()`, `geom_line()` and `geom_point()`. Use `geom_vline()` to add a vertical line for 2001. The significance of this year is that the General Assembly modified the death penalty statute in several ways, including giving prosecutors the discretion to *not* seek the death penalty in cases where it previously would have been mandatory. Format and label your plot in such a way that the information it conveys would be clear to the reader (i.e., the plot is reasonably stand-alone).

```{r}
ggplot(deathsentencesNC, aes(x = year, y = sentences)) +
geom_area(fill = "salmon") +
geom_line() +
geom_point() +
geom_vline(xintercept = 2001) + 
labs(title = "Death Sentences in NC Since 1976",
    x = "Year",
    y = "Number of Sentences",
    caption = "Vertical line represents the modification of the death penalty in 2001")
```

### Information Conveyed by Your Plots

List two pieces of information you can learn or comparisons you can make from each plot above.

#### Plot #1

* Item 1: We can learn from this plot that there are three states where the death penalty is in flux.
* Item 2: We can learn from this plot that of the remaining states half are for the death penalty and the other half are against it

#### Plot #2

* Item 1: We can learn from this plot that the Midwest is split about half-and-half in their views on the death penalty.
* Item 2: We can learn from this plot that the Northeast is mostly against the death penalty with a small percentage that says yes, in moratorium.

#### Plot #3

* Item 1: We can learn from this plot that it shows us the number of prisoners on death row and we can see that it peaks around the year 2000. 
* Item 2: We can learn from this plot that the total number of prisoners on death row has steadily decreased since the year 2000

#### Plot #4

* Item 1: We can learn from this plot that the extreme outlier is the state Texas.
* Item 2: We can learn from this plot that the data is very skewed to the right because there are a lot of states that have very little executions.

#### Plot #5

* Item 1: We can learn from this plot that there is a peak in the data between the years 1990 and 2000.
* Item 2: We can learn from this plot that the number of sentences have decreased since the year 2001, however the decrease is not constant.


<hr>

```{r}
sessionInfo()
```
